Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\

### 2025-05-25 paper.tex
checkpoint

\author{\IEEEauthorblockN{Muhammad Zain Ali}
    \IEEEauthorblockA{\textit{Dept. of Physics} \\
        \textit{Quaid-i-Azam University}\\
        Islamabad, Pakistan \\
        zain@betelwise.com}
    \and
    \IEEEauthorblockN{Khabab Nazir}
    \IEEEauthorblockA{\textit{Dept. of Physics} \\
        \textit{Quaid-i-Azam University}\\
        Islamabad, Pakistan \\
        khababnazir10@gmail.com}
}

@@

\begin{abstract}
    Planetary seismology faces significant challenges in transmitting continuous seismic data due to bandwidth and power
    constraints. Traditional event detection methods generate numerous false positives, leading to inefficient data
    transmission. This paper presents Asteroseismology, an AI-enhanced seismic event detection framework that integrates
    classical phase-picking algorithms with machine learning to optimize seismic data processing for space missions. The
    workflow begins with automatic bandpass filtering, outlier removal, and normalization to enhance signal clarity. A
    Short-Term Average to Long-Term Average (STA/LTA) analysis is then applied to detect candidate events, followed by a
    filtering mechanism that refines initial picks based on characteristic function slopes. The final step employs a
    Convolutional Neural Network (CNN), trained on seismic data from the Apollo Lunar Surface Experiments Package
    (ALSEP) and NASA’s Mars InSight mission, to distinguish true seismic events from noise, ensuring high precision
    while minimizing false detections. Designed for computational efficiency, the system processes a month’s worth of
    seismic data in under 60 seconds on an average processor. It also features six tunable parameters, allowing
    adaptation to different planetary environments and mission constraints. Initial results demonstrate that the CNN,
    despite limited training, achieves over 80 percent event detection accuracy with a false positive rate of
    approximately 5 percent. Future enhancements include refining the CNN within an Auxiliary Classifier Conditional GAN
    (AC-GAN) framework to further improve detection reliability. This AI-driven approach enables autonomous seismic data
    analysis onboard spacecraft, significantly reducing the need for raw data transmission and paving the way for more
    efficient planetary and lunar seismology missions.
\end{abstract}

@@

    A fundamental challenge in planetary seismology is the efficient management and transmission of the vast quantities
    of continuous seismic data generated by lander-based instruments. Deep-space missions operate under severe
    constraints on power, bandwidth, and data volume (Lorenz, 2015, as cited in Civilini et al., 2021; Lognonné et al.,
    2019). For instance, the SEIS instrument on the Mars InSight mission, while typically downlinking continuous data at
    2 samples per second (sps), generates a significant data volume; \cite{Lognonne2019} note that its full data
    generation capability is orders of magnitude larger than its telemetry allocation. \cite{Civilini2021} quantified
    that the telemetry for SEIS consumes approximately 1.5\% of the InSight lander's total power output. If we were to
    consider data rates comparable to those from some Apollo experiments, which utilized sampling rates such as 6.625 Hz
    for the Passive Seismic Experiment (PSE) mid-period instruments \cite{Nunn2020}, the data volume and consequently
    the power required for transmission by a modern system like SEIS would consume an even more substantial portion of a
    lander's available power. Extrapolating further, \cite{Civilini2021} projected that a similar seismic setup for a
    mission to Europa could demand 8.8–13.2\% of the lander's power solely for seismic data transmission. These figures
    underscore the critical necessity for sophisticated on-board data processing and event detection algorithms to
    prioritize scientifically valuable data segments for downlink, thereby making the most efficient use of limited
    mission resources.

@@

    This paper introduces "Asteroseismology," an AI-enhanced seismic event detection framework specifically designed to
    address the data optimization needs of extra-terrestrial missions. Our framework uniquely integrates a pipeline of
    classical phase-picking algorithms with a lightweight CNN. This hybrid approach begins with conventional signal
    conditioning, including automatic bandpass filtering, outlier removal, and normalization. Candidate seismic events
    are then identified using STA/LTA analysis, followed by a novel filtering stage based on characteristic function
    slopes to refine these initial picks. The core innovation lies in then feeding these refined candidate events,
    represented as 1D waveform segments augmented with auxiliary statistical features, into a CNN. This CNN, trained on
    a diverse dataset including lunar data from the Apollo missions and Martian data from the InSight mission, is tasked
    with the final discrimination between true seismic events and noise. By leveraging the efficiency of classical
    algorithms for initial data reduction and candidate selection, our system aims to reduce the computational burden on
    the CNN, allowing it to perform high-precision classification. Asteroseismology is designed for computational
    efficiency and adaptability through tunable parameters, making it a strong candidate for future onboard autonomous
    data processing, thereby significantly enhancing the scientific return of planetary seismology missions by
    optimizing precious data transmission resources.

@@

            The effect of applying the selected adaptive bandpass filter is shown in ``Fig.~\ref{fig:adaptive_filtering_comparison}''
            , which compares the raw seismogram with the seismogram after the optimal bandpass filter has been applied, 
            enhancing the clarity of potential signals within the chosen frequency band.
            % TODO: tell that which example event it is
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig2_adaptive_bandpass.pdf}}
                \caption{Comparison of a raw seismogram (top panel) with the same seismogram after the application 
                of the adaptively selected bandpass filter (bottom panel). The filtering aims to enhance signal 
                components within the optimal frequency range.}
                \label{fig:adaptive_filtering_comparison} 
            \end{figure}

@@

        \begin{figure}[htbp]
            \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig3_mitigation.pdf}}
            \caption{Effect of outlier mitigation. The top panel shows the seismogram after bandpass filtering, 
            potentially containing high-amplitude outliers. The bottom panel displays the same seismogram after 
            the outlier clipping process, where extreme values have been attenuated.}
            \label{fig:outlier_clipping_comparison}
        \end{figure}

@@

        \subsection{Datasets and Input Seismogram Preparation}
            \subsubsection{Data Sources (Lunar and Martian)}
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings from the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
                % \subsubsection{Waveform Segmentation and Initial Processing}
        \subsection{Performance Results}
        Initial evaluations of the framework demonstrate promising results. The CNN component, even with what is
        currently considered limited training data and refinement, achieved a **detection accuracy of over 80\%** on the
        test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize
        across different planetary environments and event characteristics.

        Crucially, the hybrid nature of the system, particularly the pre-filtering by conventional algorithms,
        contributed to a **false positive rate of approximately 5\%**. This is a significant improvement over
        traditional methods like standalone STA/LTA, which often suffer from much higher FPRs in noisy planetary
        settings. A low FPR is vital for minimizing the transmission of non-scientific data.

@@

>>>>

% \author{\IEEEauthorblockN{Muhammad Zain Ali}
%     \IEEEauthorblockA{\textit{Dept. of Physics} \\
%         \textit{Quaid-i-Azam University}\\
%         Islamabad, Pakistan \\
%         zain@betelwise.com}
%     \and
%     \IEEEauthorblockN{Khabab Nazir}
%     \IEEEauthorblockA{\textit{Dept. of Physics} \\
%         \textit{Quaid-i-Azam University}\\
%         Islamabad, Pakistan \\
%         khababnazir10@gmail.com}
% }

@@

\begin{abstract}
Planetary seismology is crucial for understanding the internal structure and evolution of celestial bodies, but missions
face severe data transmission limitations due to bandwidth and power constraints. Traditional onboard event detection
methods often struggle with high false positive rates, leading to inefficient use of these limited resources. This paper
introduces "Asteroseismology," an AI-enhanced hybrid framework designed to optimize seismic event detection for space
missions. The framework first employs an Initial Processing and Candidate Selection (IPCS) stage, which utilizes
conventional seismic algorithms. This includes adaptive bandpass filtering based on signal characteristics (power
spectrum or standard deviation), outlier removal, normalization, and STA/LTA analysis for initial event triggering. A
key refinement within this stage involves analyzing characteristic function slopes to reduce false STA/LTA detections.
Candidate events from the IPCS stage are then passed to a Machine Learning-based Event Verification (MLEV) stage. Unlike
approaches that primarily use 2D spectrograms, our MLEV stage utilizes 1D seismogram segments augmented with explicitly
engineered auxiliary statistical features (standard deviations around the arrival time) as input to a lightweight
Convolutional Neural Network (CNN). This CNN, trained on a combined dataset of seismic events and noise from both the
Apollo ALSEP (lunar) and NASA Mars InSight (Martian) missions, performs the final classification of true seismic events
versus noise. The Asteroseismology system is designed for computational efficiency, processing a month of seismic data
in under 60 seconds on an average processor, and features tunable parameters for adaptation to diverse planetary
environments. Preliminary results indicate that the CNN achieves over 80\% event detection accuracy with a false positive
rate around 5\%. Future work includes exploring advanced ML architectures like GANs for data augmentation and
physics-informed neural networks for enhanced robustness. This AI-driven hybrid approach offers a significant step
towards enabling autonomous, efficient, and accurate seismic data analysis onboard spacecraft, thereby maximizing the
scientific return from planetary seismology missions.
\end{abstract}

@@

    A fundamental challenge in planetary seismology is the efficient management and transmission of the vast quantities
    of continuous seismic data generated by lander-based instruments. Deep-space missions operate under severe
    constraints on power, bandwidth, and data volume  (Lorenz, 2015, as cited by \cite{Civilini2021};
    \cite{Lognonne2019}). For instance, the SEIS instrument on the Mars InSight mission, while typically downlinking
    continuous data at 2 samples per second (sps), generates a significant data volume; \cite{Lognonne2019} note that
    its full data generation capability is orders of magnitude larger than its telemetry allocation. \cite{Civilini2021}
    quantified that the telemetry for SEIS consumes approximately 1.5\% of the InSight lander's total power output. If
    we were to consider data rates comparable to those from some Apollo experiments, which utilized sampling rates such
    as 6.625 Hz for the Passive Seismic Experiment (PSE) mid-period instruments \cite{Nunn2020}, the data volume and
    consequently the power required for transmission by a modern system like SEIS would consume an even more substantial
    portion of a lander's available power. Extrapolating further, \cite{Civilini2021} projected that a similar seismic
    setup for a mission to Europa could demand 8.8–13.2\% of the lander's power solely for seismic data transmission.
    These figures underscore the critical necessity for sophisticated on-board data processing and event detection
    algorithms to prioritize scientifically valuable data segments for downlink, thereby making the most efficient use
    of limited mission resources.

@@

    This paper introduces an AI-enhanced seismic event detection framework specifically designed to
    address the data optimization needs of extra-terrestrial missions. Our framework uniquely integrates a pipeline of
    classical phase-picking algorithms with a CNN. This hybrid approach begins with conventional signal
    conditioning, including automatic bandpass filtering, outlier removal, and normalization. Candidate seismic events
    are then identified using STA/LTA analysis. The core innovation lies in then feeding these refined candidate events,
    represented as 1D waveform segments augmented with auxiliary statistical features, into a CNN. This CNN, trained on
    a diverse dataset including lunar data from the Apollo missions and Martian data from the InSight mission, is tasked
    with the final discrimination between true seismic events and noise. By leveraging the efficiency of classical
    algorithms for initial data reduction and candidate selection, our system aims to reduce the computational burden on
    the CNN, allowing it to perform high-precision classification. Asteroseismology is designed for computational
    efficiency and adaptability through tunable parameters, making it a strong candidate for future onboard autonomous
    data processing, thereby significantly enhancing the scientific return of planetary seismology missions by
    optimizing precious data transmission resources.

@@

            The effect of applying the selected adaptive bandpass filter is shown in ``Fig.~\ref{fig:adaptive_filtering_comparison}''
            , which compares the raw seismogram with the seismogram after the optimal bandpass filter has been applied, 
            enhancing the clarity of potential signals within the chosen frequency band.
            % TODO: tell that which example event it is
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig2_adaptive_bandpass.pdf}}
                \caption{Comparison of a raw seismogram of an event from mars (a) with the same seismogram after the application 
                of the adaptively selected bandpass filter (b). The filtering aims to enhance signal 
                components within the optimal frequency range.}
                \label{fig:adaptive_filtering_comparison} 
            \end{figure}

@@

        \begin{figure}[htbp]
            \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig3_mitigation.pdf}}
            \caption{Effect of outlier mitigation. (a) shows the seismogram after bandpass filtering, 
            potentially containing high-amplitude outliers. (b) displays the same seismogram after 
            the outlier clipping process, where extreme values have been attenuated.}
            \label{fig:outlier_clipping_comparison}
        \end{figure}

@@

        \subsection{Datasets and Input Seismogram Preparation}
            \subsubsection{Data Sources (Lunar and Martian)}
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings from the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
                % \subsubsection{Waveform Segmentation and Initial Processing}
        \subsection{Performance Results}
        Initial evaluations of the framework demonstrate promising results. The CNN component, even with what is
        currently considered limited training data and refinement, achieved a detection accuracy of over 80\% on the
        test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize
        across different planetary environments and event characteristics.

        Crucially, the hybrid nature of the system, particularly the pre-filtering by conventional algorithms,
        contributed to a false positive rate of approximately 5\%. This is a significant improvement over
        traditional methods like standalone STA/LTA, which often suffer from much higher False Positive rate in noisy planetary
        settings. A low FPR is vital for minimizing the transmission of non-scientific data.

@@

    \bibitem{Lorenz2015} R. D. Lorenz, ``Energy cost of acquiring and transmitting science data on deep-space missions,'' \emph{J. Spacecr. Rockets}, vol. 52, no. 6, pp. 1693--1697, Nov. 2015, doi: 10.2514/1.A33298.
